Forecasting hourly hospital admission

### 2021-11-09 paper/paper.Rmd

---
title:  Forecasting short-term hourly Emergency Departement arrivals
author:
  - name: Author1
    email: email1@example.com
    affiliation: University1
    footnote: 1
  - name: Jethro Browell
    email: jethro.browell@glasgow.ac.uk
    affiliation: School of Mathematics and Statistics, University of Glasgow, UK
    footnote: 2
  - name: Ivan Svetunkov
    email: i.svetunkov@lancaster.ac.uk
    affiliation: Centre for Marketing Analytics and Forecasting, Lancaster University, UK
    footnote: 2
address:
  - code: University1
    address: Cardiff business school, 3 Colum Drive, CF10 3EU, Cardiff
  - code: University2
    address: adress2
  - code: University3
    address: adress3
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"

@@

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, cache=FALSE, fig.pos = 'H', message = FALSE, warning = FALSE)
# Make sure you have the latest versions of rmarkdown and rticle
library(tidyverse)
library(lubridate)
library(bookdown)
library(patchwork)
library(tsibble)
library(kableExtra)
library(ggplot2)
library(RColorBrewer)
library(ggridges)
library(hrbrthemes)
library(ggthemes)
library(scales)
library(ggtext)
library(extrafont)
```

@@

Forecasting Emergency Department (ED) arrivals is a critical input to inform staffing and scheduling decisions to meet the needs of patients. An accurate ED demand forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services. This is one of the best ways to optimize staffing utilization and related costs. Most current practice to optimise personnel scheduling follows the general approach originally presented in @vile2016time, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts. An accurate demand forecasting is crucial in ED services to depict various courses of action that can result in massive savings in terms of patient lives. Inability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow [@derlet2002overcrowding]. Also, it is related with increasing length of stay [@muhammet2015forecasting], low patient satisfaction, number of patients left the hospital without seen by staff, unexpected return visits to services, increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing [@rostami2020anticipating].

There exists a large literature on forecasting ED arrivals that use various methods to forecast annual[@tandberg1994time] monthly,[@chen2011long; @mai2015predicting]. daily[@rostami2020anticipating;@park2019144] and subdaily [@schweigler2009forecasting;@cheng2021forecasting] arrivals. In this paper, we only focus on forecasting hourly arrivals. In comparison to lower frequency time series forecasting such as monthly, quarterly and yearly, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
<!-- In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services -->

An accurate demand forecast by hour of the day enables planners to match staff to meet anticipated patients, reconfigure units and redeploy staff. This will have many advantages for both patients, staff and the quality of provided services. Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day. This involves the short-term decision making related to the execution of the delivery process in ED.
<!-- for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential.  -->
The combination of an hourly forecast demand, current staff being occupied, resource availability and waiting times at ED, provide information on the state of the unscheduled care system across the service. Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients. Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during  a day or a week. Staff capacities may be adjusted based on the predicted demand fluctuations by using part-time, on-call nurses, staff overtime and voluntary absenteeism. That may also require to adjust the scheduling.

There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics , climate factors , holidays. These studies use multiple approaches including Exponential Smoothing [REF], ARIMA families [], ARCH [REF], Vector Autoregressive [REF],  TBATS [] and ANN[]. Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly. They may also express nonstationarity and their profile may change over time. Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions. This is currently missing in the literature and we fill this gap by examining various forecasting models capable of considering these features. All these studies only generate the estimated future arrivals as a point forecast (a single number), which does not include any uncertainty around the number of arrivals. However, reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric. Therefore, probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing. This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than lower service levels if staff levels are insufficient. The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools. In this paper, we produce and evaluate the whole forecast distribution of hourly ED arrivals, using all forecasting models, which could be used as a risk management tool for planners and decision makers. Moreover, datasets used in some of these studies are relatively small, short length (e.g. time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable. Moreover, all previous publications  covered in this paper fails reproducibility principles.

@@

1. We develop a novel methodology to forecast short-term hourly hospital admission using the family of Generalised Additive Models that accounts for i) ..., ii) ... iii) , ... iv) (Jethro to complete)

2. We produce probabilistic forecast, in addition to the point estimation, that quantify uncertainties in future hospital admission and we evaluate its accuracy using ...;

@@

4. We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks. The proposed approach could also be generalized to forecast hourly requirements in other services such as the number of incidents or call volumes in clinical desk services.

@@

# Research background: hourly ED forecasting {#lit}

There are many studies that employs models to forecast the Emergency Department needs such as admission and arrivals. The time granularity considered by these studies spans from hourly to yearly. However, given the focus of the paper, we only focus on studies on hourly ED forecasting. Table \@ref(tab:summarylit) summarize studies in hourly forecasting in the emergency department.

```{r summarylit, cache=FALSE}
li <- readxl::read_xlsx("../table/table.xlsx")
knitr::kable(li, booktabs = T, linesep = "",caption = "Summary of studies in hourly Emergency Department forecasting") %>% kableExtra::kable_styling(latex_options = c("scale_down","hold_position"), font_size =11) %>% column_spec(3,"6em") %>%
  column_spec(4,"4em") %>% column_spec(6,"15em")  %>% column_spec(7,"4em") %>%  kableExtra::landscape()
```

@@

Predicting the arrivals of an ED future arrivals is also studied by @choudhury2020forecasting. ARIMA, Holt-Winters, TBATS, and neutral network methods were implemented to forecast hourly accident and emergency department arrivals. ARIMA model was selected as the best fit model. Authors claimed that ARIMA has provided high and acceptable hourly ED forecasting accuracy. @hertzum2017forecasting work was mentioned in this paper. This result is surprising given the existence of multiple seasonality in the hourly dataset.

@morzuch2006forecasting used the Unobserved Components Model (UCM), by which each component of the time series is separately modeled as stochastic. Double-seasonal exponential smoothing and standard Holt-Winters are used to forecast ED arrival for an horizon of 168 hours. The hourly data collected from an ED in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles. The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.

@mccarthy2008challenge employed a Poisson log-linear regression model, including independent variables such as temporal factors (i.e., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation) to forecast patient demand for ED services. Hourly arrival data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future. Authors . They present the prediction interval accuracy of $50\%$ and $90\%$ intervals for the number of hourly arrivals under the Poisson assumptions. They show that the most important predictor is hour of the day and autocorrelation lag 1. However, these findings are limited to the short number of observations (only one year of historical data).

@@

According to the studies mentioned earlier, it can be said that the existing studies have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established. Some of the studies said that the accuracy of hourly accident and emergency department forecasting model is low compared to other longer forecasting intervals like daily forecast [@boyle2012predicting; @hertzum2017forecasting]. However, some studies mentioned that the accuracy of ED hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].

There are few limitations in the literature which encourage us to undertake this research and propose examine different forecasting approaches to deal with them. These limitations are summarised as follows : (i) Current approaches to forecast hourly ED arrivals do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time; (ii) Almost all research studies produce point forecast and consequently report only point forecast accuracy. There is a lack of studies presenting probabilistic forecast of hourly ED arrivals that better represent uncertainty of future admissions, providing an holistic picture of future demand for a planner; (iii) most studies are not reproducible, as it is almost impossible to reproduce forecasting models and results from the studies; (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and (v) some studies in this area lack a rigorous experimental design, i.e. there is no benchmark method nor is forecast accuracy reported.

@@

## data {#data}

Data used in the study comprised counts of patients’ arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital. We aggregated the patients’ arrival times to obtain hourly arrivals, which is used for empirical evaluation in this study.

Figure \@ref(fig:hourly-plot-ridge) illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy but it reveals some systematic structures.

It is clear that the number of arrivals has a sub-daily structure. The ED arrivals decreases between mid-night and early morning and then increases until the evening and then decreases again. It is also clear that ED service gets systematically more visits on Monday between 8 a.m. and 5 p.m. Moreover, The number of arrivals around mid-night is  slightly higher for Saturday and Sunday.

@@

Figure \@ref(fig:hourly-plot) also highlights that there are many ouliers for almost every hour of the day that may affec the accuracy of forecasting methods, some of these outliers are might be related to holidays and special events. It is also clear that arrivals are less volatile between mid-night and early morning.

We propose a forecasting model that accounts for the systematic structure of the data.

```{r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',out.width= "70%"}

@@

As our data are counts, the natural starting point is the Poisson distribution, given by

\begin{equation}
  F_t(y_t,\lambda_t) =  \frac{\Gamma(\lfloor y_t + 1  \rfloor,\lambda_t)}{\lfloor y_t \rfloor}
  (\#eq:poissonreg)
\end{equation}

where we consider an additive model for $\lambda_t$ of the form

\begin{equation}
  \log(\lambda_t) = \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad .
 (\#eq:additivemodel)
\end{equation}

The functions $H(t)$, $D(t)$ and $Y(t)$ return the hour of the day (1--24), day of the week (1--7), and day of the year (1-366) at time $t$, respectively, and $C_t$ is the temperature at time $t$.

Truncated Normal...

\begin{equation}
  F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
 (\#eq:truncatedn)
\end{equation}

with

\begin{align*}
  \mu_t = &  \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
  log(\sigma_t) & = f(H(t)) .
\end{align*}

Truncated $t$ distribution...

Negative binomial...

@@

Finally, we have calculated the computational time for one iteration on the first rolling origin to compare the speed of each function. All functions were re-estimated on each iteration. ADAM and Poisson regression estimated the parameters taking as the pre-initials the ones obtained in the initial model application to the data in the first origin. This allowed to speed up the computation for these two models. The initial estimation of ADAM took approximately one hour and 25 minutes. Each step in the experiment took the time shown below.

## Case study

The data is portioned into training and test data, with all model development and hyper-parameter tuning performed using training data only. Data from 2014-04-01 to 2018-02-28 are used for training, and from 2018-03-01 to 2019-02-28 for testing.

@@

@@

Probabilistic forecasts are evaluated following the principle of *sharpness subject to calibration*, meaning that the sharper forecast is prefered provided that it is calibrated. Mis-calibrated forecasts are unsuitable for use in decision-making so should be excluded. Calibration is evaluated visually in \@ref(quantile-bias), which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal and $t$ family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and iETS models showing good calibration across most probability levels. Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.

```{r Pinball, fig.cap= "pinball ...", fig.align='center',out.width= "90%"}
```

```{r lead-time-rel, fig.cap= "pinball ...", fig.align='center',out.width= "90%"}
```

```{r Reliability, fig.cap= "Reliability ...", fig.align='center',out.width= "90%"}
```

```{r quantile-bias, fig.cap= "Quantile ...", fig.align='center',out.width= "90%"}
```

```{r tab-results, out.width= "50%"}
results_table <- read_rds("results_table.rds")
knitr::kable(results_table, booktabs = T, 
             linesep = "",caption = "Summary of studies in hourly emergency care forecasting") %>% kableExtra::kable_styling(latex_options = c("hold_position"), font_size =11)
```

Evaluation metrics from the test period are presented in \@ref(tab-results) which are ordered by Quantile Bias. The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecast with Quantile Biases of 0.037 and above, ETS(XXX) being the only exception with a value of 0.019.

>>>>

---
title:  Forecasting hourly hospital admission
author:
  - name: Author1
    email: email1@example.com
    affiliation: University1
    footnote: 1
  - name: Jethro Browell
    email: jethro.browell@glasgow.ac.uk
    affiliation: School of Mathematics and Statistics, University of Glasgow, UK
    footnote: 2
  - name: Ivan Svetunkov
    email: i.svetunkov@lancaster.ac.uk
    affiliation: Centre for Marketing Analytics and Forecasting, Lancaster University, UK
    footnote: 2
address:
  - code: University1
    address: Cardiff business school, 3 Colum Drive, CF10 3EU, Cardiff
  - code: University2
    address: adress2
  - code: University3
    address: adress3
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"

@@

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, cache=TRUE, fig.pos = 'H', message = FALSE, warning = FALSE)
# Make sure you have the latest versions of rmarkdown and rticle
library(tidyverse)
library(lubridate)
library(bookdown)
library(patchwork)
library(tsibble)
library(kableExtra)
library(ggplot2)
library(data.table)
library(RColorBrewer)
library(ggridges)
library(hrbrthemes)
library(ggthemes)
library(scales)
library(ggtext)
library(extrafont)
```

@@

Forecasting hospital admission is a critical input to inform staffing and scheduling in hospital services to meet the needs of patients. An accurate hospital admission forecast contributes to a better decision making process regarding the resources needed to provide services to the number and type of patients requiring hospital services. This is one of the best ways to optimize staffing utilization and related costs. Most current practice to optimise personnel scheduling follows the general approach originally presented in @vile2016time, which recommends that the following steps be taken to roster employees: (i) forecast demand; (ii) convert demand forecasts into staffing requirements; (iii) schedule shifts optimally; and (iv) assign employees to shifts. An accurate demand forecasting is crucial in Accident & Emergency and Ambulance Services in particular, to depict various courses of action that can result in massive savings in terms of patient lives [@rostami2020anticipating].

Unability to match the staff with the demand results in patient overcrowding in the system which is a serious problem that causes challenging situations on patient flow [@derlet2002overcrowding]. Also, it is related with increasing length of stay [@muhammet2015forecasting], low patient satisfaction, number of patients left the hospital without seen by staff, unexpected return visits to services, increasing health care costs, inaccuracy in electronic medical record, and reported waiting times without incurring last-minute expenses, such as overtime or supplemental staffing.

There exists a large literature on forecasting hospital admission - including emergency department and ambulance services - that use various methods to forecast annual[@tandberg1994time] monthly,[@chen2011long; @mai2015predicting]. daily[@rostami2020anticipating;@park2019144] and subdaily [@schweigler2009forecasting;@cheng2021forecasting] admission. In this paper, we only focus on forecasting hourly hospital admission. In comparison to lower frequency time series forecasting such as monthly quarterly and yearly admission, hourly forecasts are challenging because the noise caused by random variation may overshadow any pattern in the time series.
<!-- In this respect, forecasts of daily or monthly arrivals are likely easier but target decisions about staff allocation and the like, not the ongoing scheduling and rescheduling of how the available resources are divided among the patients in need of emergency services --> An accurate forecast of the admission demand by hour of the day enables managers to match staff to meet anticipated patients, reconfigure units and redeploy staff. This will have many advantages for both patients, staff and the quality of provided services.  Hourly forecasts are required to inform the short-term operational planning for the current and the upcoming shifts of the day. This involves the short-term decision making related to the execution of the delivery process for various health care services such as Ambulatory, Emergency, Surgical, Inpatient, Home and Residential. The combination of an hourly forecast demand, current staff being occupied (e.g incidents being attended), resource availability and waiting times at hospitals, provide information on the state of the unscheduled care system across the service. Having this full picture enables the delivery managers to focus on the areas that require intervention to enable the most effective delivery of the service to the patients. Moreover, there are often unplanned events, such as walk-in patients, extended consultation times, during  a day or a week. Staff capacities may be adjusted based on the predicted demand fluctuations by using part-time, on-call nurses, staff overtime and voluntary absenteeism. That may also require to adjust the scheduling.

There are few studies that look at forecasting hourly hospital admission using historical time series data and/or predictors such as patient characteristics , climate factors , holidays. These studies use multiple approaches including Exponential Smoothing [REF], ARIMA families [], ARCH [REF], Vector Autoregressive [REF],  TBATS [] and ANN[]. Hourly time series generally exhibit multiple seasonal cycles of different lengths such as hourly, daily, weekly and yearly. They may also express nonstationarity and their profile may change over time. Therefore, an appropriate forecasting model should take these features into account, to accurately predict hourly demand admissions. This is currently missing in the literature and we fill this gap by examining various forecasting models capable of considering these features. All these studies only generate the estimated future admission as a point forecast (a single number), which does not report any uncertainty around the admission. Reporting uncertainty is critical because the consequences of imperfect staffing are asymmetric. Therefore, probabilistic forecasts are necessary to make decisions that balance the cost associated with under and over staffing. This asymmetry arises because it is preferable to incur a small opportunity cost associated under utilised staff rather than lower service levels if staff levels are insufficient. The lack of probabilistic forecast is one of the main bottlenecks in the deployment of generated forecasts in the staffing and planning tools. In this paper, we produce and evaluate the whole forecast distribution of hourly hospital admission, using all forecasting models, which could be used as a risk management tool for planners and decision makers. Aditionally, datasets used in some of these studies are relatively small, short length (e.g. time period of 1 year), which make it challenging to report the forecast accuracy using robust approaches such as time series cross validation and such results might not be reliable. Moreover, all previous publications  covered in this paper fails reproducibility principles (Boylan, 2016).

@@

1. We produce probabilistic forecasts, in addition to the point estimation, which quantify uncertainties in future hospital admission, and compare different forecasting methods using a suite of established evaluation metrics;

2. We develop a novel approach to forecast day-ahead hourly hospital admission using Generalised Additive Models for Location Scale and Shape, which accounts for i) the bounded and non-Gaussian distribution of admissions, ii) multiple seasonalities, weather and holiday effects, and  iii) variation in forecast uncertainty;

@@

4. We provide data and code enabling reproduction and refinement of the proposed approach and benchmarks. The proposed approach could also be generalised to forecast hourly requirements in other services such as the number of incidents or call volums in clinical desk services.

@@

# Research background: hourly hospital admission forecasting {#lit}

Other papers to review and possibly cite:
- Daily visits using Prophet @park2019144
- Daily using internet data @ekstrom2015forecasting
- Review paper: @wargon2009systematic
- Others: @boyle2012predicting, 
- Impact of weather @parsons2011modelling, @macgregor2003effect

hourly:
@cheng2021forecasting
@rocha2021forecasting

Table \@ref(tab:summarylit) summarise studies in hourly forecasting in emergency and urgent care.

```{r summarylit}
li <- readxl::read_xlsx("../table/table.xlsx")
knitr::kable(li, booktabs = T, linesep = "",caption = "Summary of studies in hourly emergency care forecasting") %>% kableExtra::kable_styling(latex_options = c("scale_down","hold_position"), font_size =11) %>% column_spec(3,"6em") %>%
  column_spec(4,"4em") %>% column_spec(6,"15em")  %>% column_spec(7,"4em") 
```

@@

Predicting the arrivals of accident and emergency department future patients is studied by @choudhury2020forecasting ARIMA, Holt-Winters, TBATS, and neutral network methods were implemented to forecast hourly accident and emergency department arrivals. ARIMA model was selected as the best fit model and it has provided high and acceptable hourly accident and emergency department forecasting accuracy. @hertzum2017forecasting work was mentioned in this paper. It is said that residual normality, stationarity, and autocorrelation have not been tested in @hertzum2017forecasting paper and this might be the cause of accuracy problems. However, residual normality, stationarity, and autocorrelation are tested and compared with Holt-Winters, TBATS, and neutral network methods in @choudhury2020forecasting 
@morzuch2006forecasting use the Unobserved Components Model (UCM), by which each component of the time series is separately modelled as stochastic; double-seasonal exponential smoothing and standard Holt-Winters to forecast ED arrival for an horizon of 168 hours. The hourly data collected from an Emergency department in Pennsylvania showed no trend, and two seasonal cycles: a within-day and a within-week seasonal cycles. The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly seasonality of the time series.

@mccarthy2008challenge predict patient demand for ED services by characterising ED arrivals. Hourly arrival data of ED arrivals in the 1-year study period was deployed to forecast from 1 hour to 24 hours into the future. Authors use a Poisson log-linear regression model, including independent variables such as temporal factors (i.e., hour-of-day, day- of-week, type-of-day, season, and calendar year), patient characteristics (i.e., age, gender, insurance status, triage level, mode of arrival, and ambulance diversion status) and climatic factors (i.e., temperature and precipitation).
Then, they presente the predictiona interval accuracy of 50 and $90\%$ intervals for the number of hourly arrivals under the Poisson assumptions. They show that the most important predictor is hour of the day and autocorrelation lag 1. However, these findings are limited to the short period of data (one year).

@@

@steins2019forecasting aimed to develop a forecasting model for predicting the number of ambulance services calls per hour and geographical areas, to support managers in decisions-making and to investigate which ones were the factors that affected the number of calls. Data collected consisted of a time and location of historical ambulance call data for three counties in Sweden and a list of explanatory factors of the area, such as socioeconomic and geographic.  In order to deal with large number of zeros in the data, authors developed zero-inflated Poisson (ZIP) and zero-inflated negative binomial (ZINB) regression models. These were then compared to the currently existing forecasting system, based on moving average with seasonality weights, using ME, MAE and RMSE. Firstly, the factors affecting the number of ambulance calls were found to be the following: population in different age groups, median income, length of road, number of nightlife spots (the number of restaurants), day of the week and hour of the day. Secondly, it was found that the older population (65-100) generated more ambulance calls and that ZIP model performs better than the current model. However, the improvement provided by the more advance model was not much greater than the one provided by the existing model. Authors suggested that it could be because the population, which was used as an independent variable in both models, was so dominant compared to the other factors. 
Moreover, both models either underestimated or overestimated the number of calls. Authors suggested that the inability to capture a positive trend resulted in underestimation, and the opposite was due to a negative trend. Therefore, authors recommended that further research should add certain temporal variables able to capture the trend.

According to the studies mentioned earlier, it can be said that the existing studies have shown complications in forecasting hourly patient accident and emergency department visits and the application of forecasting hourly patients visits is not well established. Some of the studies said that the accuracy of hourly accident and emergency department forecasting model is low compared to other longer forecasting intervals like daily forecast [@boyle2012predicting; @hertzum2017forecasting]. However, some studies mentioned that the accuracy of accident and emergency department hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].

There are few limitations in the literature which encourage us to undertake this research and propose examine different forecasting approaches to deal with them. These limitations are summarised as follows : (i) Current approaches to forecast hourly hospital admission do not fully consider the feature of data such as multiple seasonal cycles and changing profile over time; (ii) Almost all research studies produce point forecast and consequently report only point forecast accuracy. There is a lack of studies presenting probabilistic forecast of hourly hospital admission that better represent uncertainty of future admissions, providing an holistic picture of future admission for a planner; (iii) most studies are not reproducible as it is almost impossible to reproduce forecasting models and results; (iv) studies are limited in terms of the length of historical data used for training purposes and forecast performance evaluation and (v) some studies in this area lack a rigorous experimental design, i.e. there is no benchmark method nor is forecast accuracy reported.

@@

We develop a range of models for forecasting hospital admissions using a multi-year dataset from a hospital in the UK. In order to facilitate model development and testing, we divide the data into a training and test period. We use cross-validation to iteratively fit, evaluate and develop models using the training data, before settling on their optimal configuration and testing them on the as-yet unseen test data. Data from 2014-04-01 to 2018-02-28 are used for training, and from 2018-03-01 to 2019-02-28 for testing.

Forecast are issued every day at midnight and mid-day for lead-times from 0 to 48 hours-ahead on a rolling basis. In order to compare both parametric and non-parametric probabilistic forecasts, we evaluate quantiles of the predictive distribution at probability level 0.05, 0.10,... 0.95, as well as the predictive expectation.

In the remainder of this section we explore the data in more detail, describe the forecast evaluation metrics we will employ, and introduce a range of benchmark models for producing forecasts. The benchmarks serve as reference methods and include some simple methods as well as more sophisticated routines designed to work `out-of-the-box', such as Facebook's Prophet. The performance of these methods provides an important reference, and any proposed novel solution should offer material benefits to justify increased complexity.

## Data exploration {#data}

The data we use in this case study are from hospital XYZ, and comprises records of arrivals at Accident and Emergency at hourly granularity. The training data runs from 01/04/2014 to 28/02/2018, and the test data runs from 1/3/2018 to 28/2/2019.

Figure \@ref(fig:hourly-plot-ridge)

@@

Figure \@ref(fig:hourly-plot)

```{r hourly-plot, fig.cap = 'Seasonal plot of ED attendance', fig.align='center',out.width= "80%"}

@@

As our data are counts, the natural starting point is the Poisson distribution, given by
\begin{equation}
F_t(y_t,\lambda_t) =  \frac{\Gamma(\lfloor y_t + 1  \rfloor,\lambda_t)}{\lfloor y_t \rfloor}
(\#eq:poissonreg)
\end{equation}
<!-- where we initially consider an additive model for $\lambda_t$ of the form -->
<!-- \begin{equation} -->
<!--   \log(\lambda_t) = \sum_{i=1}^7 \beta_i D_i(t) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad , -->
<!--  (\#eq:additivemodel) -->
<!-- \end{equation} -->
<!-- where the functions $H(t)$ and $Y(t)$ return the hour of the day (1--24) and day of the year (1-366) at time $t$, respectively, $D_i(t)$ is a dummy variable for day of the week at $t$, and $C_t$ is the temperature at time $t$. Analysis of this model fit on the training data revealed systematic biases which were identified as relating to public and school holidays. We therefore expand the model to include these effects as follows, -->
<!-- \begin{equation} -->
<!--   \log(\lambda_t) = \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad , -->
<!--  (\#eq:additivemodel_v2) -->
<!-- \end{equation} -->
<!-- where $D^{+}_i(t)$ is a dummy variable for day type including Monday to Sunday, Public Holiday, Christmas Day and New Year's Day, and $S_k(t)$ is a dummy variable for School Holidays. -->
where we consider an additive model for $\lambda_t$ of the form
\begin{equation}
\log(\lambda_t) = \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad ,
(\#eq:additivemodel_v2)
\end{equation}
where the functions $H(t)$ and $Y(t)$ return the hour of the day (1--24) and day of the year (1-366) at time $t$, respectively, where $D^{+}_i(t)$ is a dummy variable for day type including Monday to Sunday, Public Holiday, Christmas Day and New Year's Day, $S_k(t)$ is a dummy variable for School Holidays, and $C_t$ is the temperature at time $t$. Public and school holidays were included following experiments on the training data which revealed systematic biases associated with these periods.

However, experiments on the training data reveal that calibration of forecasts based on the Poisson distribution is poor, suggesting that the shape of the distribution is unsuitable for the present application. In particular, we observe that forecast uncertainty appears to vary depending on the time of day and possibly other explanatory variables. Therefore, we consider more flexible, two parameter distributions so that we are able to specify additive models for both location and scale parameters, specifically the truncated Normal distribution, with truncation at 0. The resulting density forecasts are given by
\begin{equation}
F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)}
(\#eq:truncatedn)
\end{equation}
with additive models
\begin{align*}
\mu_t = &  \sum_{i=1}^10 \beta_i D^{+}_i(t) + \sum_{j=1}^10 D^{+}_j(t) f_j(H(t)) + \sum_{k=1}^2 \beta_{10+k} S_k(t) + \sum_{l=1}^2 S_l(t) f_l(H(t)) + t + f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\
log(\sigma_t) & = \sum_{i=1}^10 D^{+}_i(t) f(H(t))
(\#eq:additivemodel_NOtr_v2)
\end{align*}
for the mean and variance parameters. We also performed experiments with the truncated $t$ distribution and negative binomial distribution, but these did not results in forecasts as well calibrated as the truncated normal.

# Result and discussion {#result}

@@

## Model fits

(Plots generated with theme_few - they look nice, good suggestion! Not sure how to include them in bookdown, can you help, Bahman? The plots are in the results folder for now, can either point to them from here, or I can change the Evauation script to save them somewhere else, and maybe in a different format? I guess PDF might be preferred. I realise that while I prepared my secript to produce point forecasts I haven't run it yet so will do this next week.)

@@

### Truncated Normal GAMLSS Model

We can support the interpretation the models given in Equation \@ref(eq:additivemodel_NOtr_v2) after the parameters have been estimated by visualising their reposes to inputs. Figure \@ref(fig-NOtr-model) shows the effect of time of day and day-type on the level and spread of predicted arrivals.

```{r fig-NOtr-model}

load("NOtr2_model.Rda")

plot_data <- data.frame(Hour = 0:23)
for(i in c("Mon","Tue","Wed","Thu","Fri","Sat","Sun","Public Holiday","Christmas Day","New Years Day")){
  pef <- gamlss::getPEF(h2_gamlss2,term="clock_hour",parameter = "mu",fixed.at = list(dow3=i))
  plot_data[[i]] <- pef(0:23)
}

plot_data <- reshape::melt.data.frame(plot_data,id.vars="Hour",variable_name="Day Type")

plot_1 <- ggplot(plot_data,aes(x=Hour,y=value,color=`Day Type`)) + geom_smooth(se = F,span=0.5) +
  ylab("Effect [Expected Arrivals]") + theme_few()

# plot_1

plot_data2 <- data.frame(Hour = 0:23)
for(i in c("Mon","Tue","Wed","Thu","Fri","Sat","Sun","Public Holiday","Christmas Day","New Years Day")){
  pef <- gamlss::getPEF(h2_gamlss2,term="clock_hour",parameter = "sigma",fixed.at = list(dow3=i))
  plot_data2[[i]] <- pef(0:23)
}

plot_data2 <- reshape::melt.data.frame(plot_data2,id.vars="Hour",variable_name="Day Type")

plot_2 <- ggplot(plot_data2,aes(x=Hour,y=value,color=`Day Type`)) + geom_smooth(se = F,span=0.5) +
  ylab("Effect [Standard Deviation of Arrivals]") + theme_few()

#plot_2

grid.arrange(plot_1, plot_2, ncol=2)

@@

Probabilistic forecasts are evaluated following the principle of *sharpness subject to calibration*, meaning that the sharper forecast is prefered provided that it is calibrated. Mis-calibrated forecasts are unsuitable for use in decision-making so should be excluded. Calibration is evaluated visually in \@ref(Reliability), which highlights a systematic negative bias accross all probability levels in many models, with only the truncated normal GAMLSS models (NOtr-2) and ADAM-iETS models showing good calibration across all probability levels. Notably, benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance.

```{r Reliability, fig.cap= "Reliability diagram for selected methods. If forecasts are calibrated, the Empirical frequency that each probability level is deceded should match the Nominal frequency, and the reliability diagram should be a straight line on the diagonal (black dashed line).", fig.align='center',fig.width= 7,fig.height=6}
```

Evaluation metrics from all Methods on the test period are presented in \@ref(tab-results) which are ordered by Quantile Bias. The five models identified above have a Quantile Bias of 0.014 or less, which is substantially lower than the next group of forecast with Quantile Biases of 0.037 and above.

```{r tab-results, out.width= "50%"}
results_table <- read_rds("results_table.rds")
results_table <- results_table[!results_table$Method%in%c("ADAM-iETSX-Ceiling","LinearRegression","Poisson-2-I","fasster","qreg-1"),]
results_table <- results_table[order(results_table$`Quantile Bias`),]

knitr::kable(results_table, booktabs = T, 
             linesep = "",caption = "Summary of studies in hourly emergency care forecasting") %>% kableExtra::kable_styling(latex_options = c("hold_position"), font_size =11)
```
